The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2025-05-23 Paper.qmd
Minor edits

```{r setup, include=FALSE}
if(!dir.exists("Figures")) dir.create("Figures")

@@

source("analysis.R")
#source("analysis2.R") # Generates many plots. Slow!
```

@@

In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. When there are conflicts of interest between children and their caregivers, children might need to use signals whose costs outweigh the benefits of additional care for those not in need, but would result in a net benefit to those with genuine needs. In doing so, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing.

@@

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to signal need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by the skeptical responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. We conceptualize infrequent and transient sadness signals as minimal signals of need [per @maynardsmith1995]. Frequent or prolonged sadness, however, is likely a form of withdrawal or depression, both of which might be effective costly signals of need when there are conflicts between the child and their caregivers [@gaffney2022; @syme2023].

@@

utila_df |>
  dplyr::select(any_of(names(utila_df[names(longform_dict[1:23])]))) |>
  setNames(longform_dict[1:23]) |>
  datasummary_skim(fun_numeric = list(N = \(x) length(na.omit(x)), Range = \(x) str_glue("{min(x, na.rm=T)}-{max(x, na.rm=T)}"), Mean = Mean, SD = SD, Histogram = function(x) ""))
```

@@

utila_df_caregiver |>
  dplyr::select(any_of(names(utila_df_caregiver[names(longform_dict_adult_cont)]))) |>
  setNames(longform_dict_adult_cont) |>
  datasummary_skim(fun_numeric = list(N = \(x) length(na.omit(x)), Range = \(x) str_glue("{min(x, na.rm=T)}-{max(x, na.rm=T)}"), Mean = Mean, SD = SD, Histogram = function(x) ""))
```

@@

```{r}
fmt_sum <- function(v){
  v <- na.omit(v)
  n <- length(v)
  thesum <- sum(v)
  thepct <- 100*signif(thesum/n, 2)
  str_glue("{thesum} children ({thepct}%)")
}
```

@@

::: {#fig-barplot-SignalFreq}
```{r}
barplot_SignalFreq + guides(fill = guide_legend(title = 'Times/month', nrow = 1, position = 'top', reverse = T))
```

>>>>

```{r setup, include=FALSE}
if (!dir.exists("Figures")) dir.create("Figures")

@@

source("analysis.R")
# source("analysis2.R") # Generates many plots. Slow!
```

@@

In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. When there are conflicts of interest between children and their caregivers, children might need to use signals whose costs outweigh the benefits of additional care for those not in need, e.g., because increased caregiver investment is less valuable to them and/or the opportunity costs of signaling are higher, but would result in a net benefit to those with genuine needs. In doing so, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing.

@@

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to signal need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by the skeptical responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. We conceptualize infrequent and transient sadness signals as minimal signals of need [per @maynardsmith1995]. Frequent or prolonged sadness, however, is likely a form of withdrawal or depression, both of which might be effective costly signals of need when there are conflicts between the child and their caregivers [for theory and evidence, see @gaffney2022; @syme2023].

@@

utila_df |>
  dplyr::select(any_of(names(utila_df[names(longform_dict[1:23])]))) |>
  setNames(longform_dict[1:23]) |>
  datasummary_skim(
    fun_numeric = list(
      N = \(x) length(na.omit(x)),
      Range = \(x) str_glue("{min(x, na.rm=T)}-{max(x, na.rm=T)}"),
      Mean = Mean,
      SD = SD,
      Histogram = function(x) ""
    )
  )
```

@@

utila_df_caregiver |>
  dplyr::select(any_of(names(utila_df_caregiver[names(
    longform_dict_adult_cont
  )]))) |>
  setNames(longform_dict_adult_cont) |>
  datasummary_skim(
    fun_numeric = list(
      N = \(x) length(na.omit(x)),
      Range = \(x) str_glue("{min(x, na.rm=T)}-{max(x, na.rm=T)}"),
      Mean = Mean,
      SD = SD,
      Histogram = function(x) ""
    )
  )
```

@@

```{r}
fmt_sum <- function(v) {
  v <- na.omit(v)
  n <- length(v)
  thesum <- sum(v)
  thepct <- 100 * signif(thesum / n, 2)
  str_glue("{thesum} children ({thepct}%)")
}
```

@@

::: {#fig-barplot-SignalFreq}
```{r}
barplot_SignalFreq +
  guides(
    fill = guide_legend(
      title = 'Times/month',
      nrow = 1,
      position = 'top',
      reverse = T
    )
  )
```
